Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-05-31 nemolite2d_perf.tex
nemolite-paper: add para describing constant loop bounds implementation

The challenge presented to the developers of scientific software by
the drive towards Exascale computing is considerable. With power
consumption becoming the overriding design constraint, CPU clock
speeds are falling and the complex, multi-purpose compute core is
being replaced by multiple, simpler cores. This philosophy can be seen
at work in the rise of so-called accelerator-based machines in the Top
500 List (http://www.top500.org/) of supercomputers: six of the
top-ten machines in the November 2016 list make use of many-core
processors (Intel Xeon Phi, NVIDIA GPU or NRCPC SW26010). Two of the
remaining four machines are IBM BlueGene/Qs, the CPU of which has
hardware support for running 64 threads.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The compiler flags used in this work.}
\label{TABLE_compiler_flags}
\centering
\begin{tabular}{l|l|l}
\hline
Compiler  &  Flags                                    & Notes \\
\hline
Gnu       & -Ofast -mtune=native -finline-limit=50000 &   \\
Intel     & -O3 -fast -fno-inline-factor -xHost       &   \\
Cray      & -O3 -O ipa5 -h wp                         &   \\
PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 & \\
\hline
\end{tabular}
\end{table}

@@

@@

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the NEMOLite2D kernels contain conditional statements. These
statements are there to check whether \textit{e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the \texttt{safe\_address}
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

\subsubsection{In-line Momentum kernel bodies into middle-layer code}
\label{sec_inline_mom}

@@

Although the Intel compiler does do in-lining when routines are in
separate source files, we have found (both here and in our previous
work~\citep{shallow_psykal}) that the number of optimisations it
performs is reduced if it first has to in-line a routine. For the
Intel-compiled Shallow code, manually inserting kernel code at the
site of the subroutine call increased performance by about 25\%.

In fact, in-lining can have a significant effect on the Intel
compiler's ability to vectorise a loop. Taking the loop that calls the
kernel for the u-component of momentum as an example, before in-lining
the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(85,7) 
  inlined into nemolite2d.f90(86,11)
 remark #15335: loop was not vectorized: 
  vectorization possible but seems 
  inefficient. 
 --- begin vector loop cost summary ---
 scalar loop cost: 1307 
 vector loop cost: 2391.000 
 estimated potential speedup: 0.540 
 lightweight vector operations: 1133 
 medium-overhead vector operations: 1 
 vectorized math library calls: 1 
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
After we have manually in-lined the kernel body, the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(97,7) 
  inlined into nemolite2d.f90(86,11)
 LOOP WAS VECTORIZED
 --- begin vector loop cost summary ---
 scalar loop cost: 1253 
 vector loop cost: 521.750 
 estimated potential speedup: 2.350 
 lightweight vector operations: 697 
 medium-overhead vector operations: 1 
 vectorized math library calls: 1 
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
Looking at the `estimated potential speedup' in the compiler output
above, it is clear that the way in which the compiler vectorises the
two versions must be very different. This conclusion is borne out by
the fact that if one persuades the compiler to vectorise the first
version (through the use of a directive) then the performance of the
resulting binary is worse than that where the loop is left
un-vectorised. In principle this could be investigated further by
looking at the assembler that the Intel compiler generates but that is
outside the scope of this work.

@@

Comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to limitations in the compiler's analysis of
the way components of Fortran derived types were being used. Each
GOcean field object, in addition to the array holding the local
section of the field, contains a pointer to a GOcean grid object. If a
kernel requires grid-related quantities (\textit{e.g.} the grid spacing)
then these are obtained by passing it a reference to the appropriate
array within the grid object. Although these grid-related quantities
are read-only within a compute kernel, if they were referenced from the
same field object as that containing an array to which the kernel writes
then the Intel compiler identified a dependency preventing
vectorisation.  This limitation was simply removed by ensuring that
all read-only quantities were accessed via field objects that were
themselves read-only for the kernel at hand.  For instance, the call
to the continuity kernel, which confused the Intel compiler, originally
looked like this:
\begin{verbatim}
do jj = 2, N, 1
 do ji = 2, M, 1

   call continuity_code(ji, jj,      &
                        ssha%data,   &
                        sshn_t%data, &
                        ...,         &
                        ssha%grid%area_t)
\end{verbatim}
where \textit{ssha} is the only field that is written to by the kernel.
We remove any potential confusion by instead obtaining the
grid-related (read-only) quantities from a field (\textit{sshn\_t} in
this case) that is only read by the kernel:
\begin{verbatim}
do jj = 2, N, 1
 do ji = 2, M, 1

   call continuity_code(ji, jj,      &
                        ssha%data,   &
                        sshn_t%data, &
                        ...,         &
                        sshn_t%grid%area_t)
\end{verbatim}

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
of the original version of the NEMOLite2D code for the range of
compilers and CPUs considered here. Unlike the Shallow
code~\citep{shallow_psykal}, the original version of NEMOLite2D has not
been optimised. Although it is still a single source file it is, in
common with NEMO itself, logically structured with separate
subroutines performing different parts of the physics within each time
step. This structuring and the heavy use of conditional statements
favour the Intel compiler which significantly out-performs both the
Gnu and Cray compilers. Only when the problem size falls out of cache
does the performance gap begin to narrow. The reason for the
performance deficit of the Cray-compiled binary comes down to (a lack
of) SIMD vectorisation, an issue that we explore below.

\begin{figure}[!t]
\centering
\includegraphics[width=\picwidth]{orig_summary}
\caption{Summary of the performance of the original version of the 
NEMOLite2D code on 
the range of compilers and CPU-based systems under consideration.}
\label{FIG_orig_perf_summary}
\end{figure}

Moving now the to the {PS}y{KA}l version of NEMOLite2D,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest {PS}y{KA}l version for each of the compiler/CPU combinations.
While the Intel compiler still produces the best-performing binary,
the Cray-compiled binary is now a very close second. In fact, the
performance of both the Gnu- and Cray-compiled {PS}y{KA}l versions is
generally significantly greater than that of their respective original
versions. We also note that best absolute performance (in terms of
grid points processed per second) with any compiler is obtained with
the $256^2$ domain.

\begin{figure}[!t]
\centering
\includegraphics[width=\picwidth]{best_psykal_summary}
\caption{Summary of the best performance achieved by any {PS}y{KA}l 
version of NEMOLite2D with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

Figure~\ref{FIG_slowdown_summary} plots the percentage difference
between the performance of the original and the best {PS}y{KA}l versions
of NEMOLite2D for each compiler/CPU combination. This shows that it is
only the Intel-compiled binary running the $64^2$ domain that is
slower with the {PS}y{KA}l version of the code (and then only by some
3\%). For all other points in the space, the optimised {PS}y{KA}l version
of the code performs better. The results for the Cray compiler are
however somewhat sqewed by the fact that it did not SIMD vectorise key
parts of the original version (see below).

@@

Looking at the results for the Gnu compiler (and the Ivy Bridge CPU)
first, all of the steps-up in performance correspond to kernel
in-lining. None of the other transformations had any effect on the
performance of the compiled code. In fact, simply in-lining the two
kernels associated with the Momentum section was enough to improve
upon the performance of the original code.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Performance (millions of points updated per second) for the
  $256^2$ case after each code transformation. Where an optimisation
  utilises compiler-specific directives then performance figures for
  the other compilers are omitted.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c}
\hline
Compiler:           & Gnu & Intel & Cray  \\
\hline
CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
\hline
Original         & 3.87 & 12.2 & 6.83 \\
Vanilla {PS}y{KA}l  & 3.07 & 6.55 & 6.73 \\
Safe-address     & --   & --   & 6.95 \\
In-line Momentum & 4.31 & 10.7 & 6.93 \\
SIMD Momentum    & --   & --   & 11.8 \\
\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
In-line Continuity         & 4.83 & 11.8 & 11.6 \\
In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
In-line field copies       & 5.92 & 12.5 & 11.4 \\
\hline
\parbox{2.5cm}{\raggedright \%-speed-up of best \textit{c.f.} original} & 34.6 & 2.39 & 42.2 \\
\hline
\end{tabular}
\end{table}

>>>>

The challenge presented to the developers of scientific software by
the drive towards Exascale computing is considerable. With power
consumption becoming the overriding design constraint, CPU clock
speeds are falling and the complex, multi-purpose compute core is
being replaced by multiple, simpler cores. This philosophy can be seen
at work in the rise of so-called accelerator-based machines in the Top
500 List of supercomputers (http://www.top500.org/): six of the
top-ten machines in the November 2016 list make use of many-core
processors (Intel Xeon Phi, NVIDIA GPU or NRCPC SW26010). Two of the
remaining four machines are IBM BlueGene/Qs, the CPU of which has
hardware support for running 64 threads.

@@

\begin{table}[!t]
\caption{The compiler flags used in this work.}
\label{TABLE_compiler_flags}
\centering
\begin{tabular}{l|l}
\hline
Compiler  &  Flags                                                       \\
\hline
Gnu       & -Ofast -mtune=native -finline-limit=50000                    \\
Intel     & -O3 -fast -fno-inline-factor -xHost                          \\
Cray      & -O3 -O ipa5 -h wp                                            \\
PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 \\
\hline
\end{tabular}
\end{table}

@@

\subsubsection{Constant loop bounds}

In the vanilla {PS}y layer, the lower and upper bounds for each loop
over grid points are obtained from the relevant components of the
derived type representing the field being updated by the kernel being
called from within the loop. In our previous
work~\citep{shallow_psykal} we found that the Cray compiler in
particular produced more performant code if we changed the PSy layer
such that the array extents are looked up once at the beginning of the
PSy routine and then used to specify the loop bounds. We have therefore
applied that transformation to the PSy layer of NEMOLite2D.

@@

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the NEMOLite2D kernels contain conditional statements. These
statements are there to check whether \textit{e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the
GOcean infrastructure to allocate all field-data arrays with extents
greater than strictly required.  We were then able to safely add the
\texttt{safe\_address} before all of the loops where the Cray compiler
indicated it might be useful (the Momentum loops and some of the BC
loops).

\subsubsection{In-line Momentum kernel bodies into middle-layer}
\label{sec_inline_mom}

@@

Although the Intel compiler does do in-lining when routines are in
separate source files, we have found (both here and in our previous
work~\citep{shallow_psykal}) that the extent of the optimisations it
performs is reduced if it first has to in-line a routine. For the
Intel-compiled Shallow code, manually inserting kernel code at the
site of the subroutine call increased performance by about 25\%.

In fact, in-lining can have a significant effect on the Intel
compiler's ability to vectorise a loop. Taking the loop that calls the
kernel for the u-component of momentum as an example, before in-lining
the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(85,7) 
  inlined into nemolite2d.f90(86,11)
 remark #15335: loop was not vectorized: 
  vectorization possible but seems 
  inefficient. 
 --- begin vector loop cost summary ---
 scalar loop cost: 1307 
 vector loop cost: 2391.000 
 estimated potential speedup: 0.540 
 ...
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
After we have manually in-lined the kernel body, the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(97,7) 
  inlined into nemolite2d.f90(86,11)
 LOOP WAS VECTORIZED
 --- begin vector loop cost summary ---
 scalar loop cost: 1253 
 vector loop cost: 521.750 
 estimated potential speedup: 2.350 
 ...
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
Looking at the `estimated potential speedup' in the compiler output
above, it is clear that the way in which the compiler vectorises the
two versions must be very different. This conclusion is borne out by
the fact that if one persuades the compiler to vectorise the first
version (through the use of a directive) then the performance of the
resulting binary is worse than that where the loop is left
un-vectorised. In principle this could be investigated further by
looking at the assembler that the Intel compiler generates but that is
outside the scope of this work.

@@

Comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to limitations in the compiler's analysis of
the way components of Fortran derived types were being used. Each
GOcean field object, in addition to the array holding the local
section of the field, contains a pointer to a GOcean grid object. If a
kernel requires grid-related quantities (\textit{e.g.} the grid spacing)
then these are obtained by passing it a reference to the appropriate
array within the grid object. Although these grid-related quantities
are read-only within a compute kernel, if they were referenced from the
same field object as that containing an array to which the kernel writes
then the Intel compiler identified a dependency preventing
vectorisation.  This limitation was simply removed by ensuring that
all read-only quantities were accessed via field objects that were
themselves read-only for the kernel at hand.  For instance, the call
to the continuity kernel, which confused the Intel compiler, originally
looked like this:
\begin{verbatim}
   call continuity_code(ji, jj,      &
                        ssha%data,   &
                        sshn_t%data, &
                        ...,         &
                        ssha%grid%area_t)
\end{verbatim}
where \textit{ssha} is the only field that is written to by the kernel.
We remove any potential confusion by instead obtaining the
grid-related (read-only) quantities from a field (\textit{sshn\_t} in
this case) that is only read by the kernel:
\begin{verbatim}
   call continuity_code(ji, jj,      &
                        ssha%data,   &
                        sshn_t%data, &
                        ...,         &
                        sshn_t%grid%area_t)
\end{verbatim}

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
of the original version of the NEMOLite2D code for the range of
compilers considered here. Unlike the Shallow
code~\citep{shallow_psykal}, the original version of NEMOLite2D has not
been optimised. Although it is still a single source file it is, in
common with NEMO itself, logically structured with separate
subroutines performing different parts of the physics within each time
step. This structuring and the heavy use of conditional statements
favour the Intel compiler which significantly out-performs both the
Gnu and Cray compilers. Only when the problem size spills out of cache
does the performance gap begin to narrow. The reason for the
performance deficit of the Cray-compiled binary comes down to (a lack
of) SIMD vectorisation, an issue that we explore below.

\begin{figure}[!t]
\centering
\includegraphics[width=\picwidth]{orig_summary}
\caption{Summary of the performance of the original version of the 
NEMOLite2D code on an Intel Ivy Bridge CPU for
the range of compilers under consideration.}
\label{FIG_orig_perf_summary}
\end{figure}

Moving now the to the {PS}y{KA}l version of NEMOLite2D,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest {PS}y{KA}l version for each of the compiler/problem-size combinations.
While the Intel compiler still produces the best-performing binary,
the Cray-compiled binary is now a very close second. In fact, the
performance of both the Gnu- and Cray-compiled {PS}y{KA}l versions is
generally significantly greater than that of their respective original
versions. We also note that best absolute performance (in terms of
grid points processed per second) with any compiler is obtained with
the $256^2$ domain.

\begin{figure}[!t]
\centering
\includegraphics[width=\picwidth]{best_psykal_summary}
\caption{Summary of the best performance achieved by any {PS}y{KA}l 
version of NEMOLite2D for each of the compilers under consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

Figure~\ref{FIG_slowdown_summary} plots the percentage difference
between the performance of the original and the best {PS}y{KA}l versions
of NEMOLite2D for each compiler/problem-size combination. This shows that it is
only the Intel-compiled binary running the $64^2$ domain that is
slower with the {PS}y{KA}l version of the code (and then only by some
3\%). For all other points in the space, the optimised {PS}y{KA}l version
of the code performs better. The results for the Cray compiler are
however somewhat sqewed by the fact that it did not SIMD vectorise key
parts of the original version (see below).

@@

Looking at the results for the Gnu compiler (and the Ivy Bridge CPU)
first, all of the steps-up in performance correspond to kernel
in-lining. None of the other transformations had any effect on the
performance of the compiled code. In fact, simply in-lining the two
kernels associated with the Momentum section was sufficient to exceed
the performance of the original code.

@@

\begin{table}[!t]
  \caption{Performance (millions of points updated per second) on an Intel
    Ivy Bridge CPU for the
  $256^2$ case after each code transformation. Where an optimisation
  utilises compiler-specific directives then performance figures for
  the other compilers are omitted.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c}
\hline
Compiler:           & Gnu & Intel & Cray  \\
\hline
Original         & 3.87 & 12.2 & 6.83 \\
Vanilla {PS}y{KA}l  & 3.07 & 6.55 & 6.73 \\
Safe-address     & --   & --   & 6.95 \\
In-line Momentum & 4.31 & 10.7 & 6.93 \\
SIMD Momentum    & --   & --   & 11.8 \\
\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
In-line Continuity         & 4.83 & 11.8 & 11.6 \\
In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
In-line field copies       & 5.92 & 12.5 & 11.4 \\
\hline
\parbox{2.5cm}{\raggedright \%-speed-up of best \textit{c.f.} original} & 34.6 & 2.39 & 42.2 \\
\hline
\end{tabular}
\end{table}
